Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-21 ms.tex
Moved 'Results' to 'VERITAS Results' and before 'Fermi' section

\altaffiltext{1}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{ Institut f\"ur Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON, M5S 3H8, Canada}
\altaffiltext{7}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and succesively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, a present day Hubble constant of $H_0 = 100 h \mbox{ km s}^{-1} \mbox{ Mpc}^{-1}$ with $h=0.7$ has been used.

@@

\section{Fermi-LAT Observations and Analysis}
The Large Area Telescope (LAT) on board the \emph{Fermi} Gamma-ray Space Telescope (\emph{Fermi}) has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The \emph{Fermi}-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is thus very complimentary to the VERITAS observations. \citet{article:Ackermann_etal:2010} reported on the search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an updated analysis which includes additional data taken during the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in \citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events calibrated  in flight) were used throughout this work. Only photons above 1 GeV are considered, where the constraints on the flux are most meaningful. The 99\% confidence level flux upper limits in the energy range 1-100 GeV, assuming both a unresolved, point-like or spatially extended emission are tabulated in Table \ref{tabel:fermi}. The gamma-ray emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$ and $2.5$.

\section{Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$ distribution is a radial comparison of the ON-source event distribution with the estimated background. It extends out to a radius of $0.42^{\circ}$ to cover all regions of interest considered in this work. A 99\% confidence level upper limit is calculated for each region of interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian background. A lower bound of zero is imposed on the signal rate to handle the negative excesses seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent of unity.

@@

@@

\begin{deluxetable}{lcccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{\emph{Fermi}-LAT upper limits for the Coma cluster core.}
\tablehead{
\colhead{Spatial model} & \multicolumn{6}{c}{Flux U.L.}
}
\startdata
Point source & 1.73 & 1.81 & 0.487 & 0.526 & 0.627 & 0.635
\enddata
\end{deluxetable}

>>>>

\altaffiltext{1}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{Institut f\"ur Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON, M5S 3H8, Canada}
\altaffiltext{7}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and succesively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, a present day Hubble constant of $H_{0} = 100h$ km s$^{-1}$ Mpc$^{-1}$ with $h=0.7$ has been used.

@@

\section{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$ distribution is a radial comparison of the ON-source event distribution with the estimated background. It extends out to a radius of $0.42^{\circ}$ to cover all regions of interest considered in this work. A 99\% confidence level upper limit is calculated for each region of interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian background. A lower bound of zero is imposed on the signal rate to handle the negative excesses seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent of unity.

@@

\section{Fermi-LAT Analysis and Results}
The Large Area Telescope (LAT) on board the \emph{Fermi} Gamma-ray Space Telescope (\emph{Fermi}) has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The \emph{Fermi}-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is thus very complimentary to the VERITAS observations. \citet{article:Ackermann_etal:2010} reported on the search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an updated analysis which includes additional data taken during the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in \citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events calibrated  in flight) were used throughout this work. Only photons above 1 GeV are considered, where the constraints on the flux are most meaningful. The 99\% confidence level flux upper limits in the energy range 1-100 GeV, assuming both a unresolved, point-like or spatially extended emission are tabulated in Table \ref{table:fermi}. The gamma-ray emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$ and $2.5$.

@@

\begin{deluxetable}{lcccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{\emph{Fermi}-LAT upper limits for the Coma cluster core.}
\tablehead{
\colhead{Spatial model} & \multicolumn{6}{c}{Flux U.L.}
}
\startdata
Point source & 1.73 & 1.81 & 0.487 & 0.526 & 0.627 & 0.635
\enddata
\label{table:fermi}
\end{deluxetable}
